Automatically Splitting a Staged Lambda Calculus

### 2015-07-09 paper/semantics.tex
said what metaprogrammings goal was

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This difference is natural, given the goals of metaprogramming and partial evaluation versus our applications.
\TODO Actually explain what MP and PE want this for.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

>>>>

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code writing code.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}
